Scaling rates of true polar wander in convecting planets and moons

### 2016-03-08 tpw_rate.tex
Change angles so as not to reuse alpha

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/reference_frames.pdf}
\caption{Relevant vectors and angles for the TPW analysis. The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively. Since the choice of geographic axes is arbitrary, we assume that at this instant the $\mathbf{E}$-frame and the geographic frames are colocated (though at future times they will not be). The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $\mathbf{e}_3$-axis. For illustration, we assume the longitude $\psi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane. True polar wander in the geographic reference frame moves the rotation axis towards the $\mathbf{e}_3$-axis, shown by $\dot{\mitbf{\omega}}$. However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$-frame. This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$ in the geographic frame, which is defined by colatitude $\alpha$ and longitude $\beta$. The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$-frame. The motion of $\mitbf{\omega}$ in the $\mathbf{E}$-frame is shown by $\dot{\mitbf{\omega}}^\prime$. Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Sections~\ref{sec:reference_frames} and \ref{sec:theta}}
\label{fig:reference_frames}
\end{figure*}

@@

There are two processes at work which control the evolution of $\theta$, its growth through perturbations in the convective moment and 
its decay via relaxation of the pole towards the maximum moment of inertia.
We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
the rotation of the body-fixed frame relative to inertial space, as shown in Equation~\eqref{eq:euler}).
The $\mathbf{E}$-frame slowly rotates with respect to the geographic frame, described by the rotation vector $\mathbf{\Psi}$ 
(here $\mitbf{\Psi}$ defines the rotation of the $\mathbf{E}$-frame in the geographic frame, see Figure~\ref{fig:reference_frames}).
We can then write the rate of change of $\mitbf{\omega}$ in the $\mathbf{E}$-frame (denoted by primes)
\begin{equation}
\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
Equation~\eqref{eq:liouville_ode} then becomes
\begin{equation}
 \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\label{eq:liouville_ode_new_frame}
\end{equation}
The advantage of expressing TPW in the $\mathbf{E}$-frame is that the entire equation
may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that $\phi=0$
(for nonzero $\phi$ the equations become more complex, but the basic physics is unchanged).
The direction of $\Psi$ is given by colatitude $\alpha$ and longitude $\beta$, as shown in Figure~\ref{fig:reference_frames}.

We may solve Equation~\eqref{eq:liouville_ode_new_frame} for $\dot{\theta}$ 
(i.e., the colatitudinal velocity in the $\mathbf{E}$ frame, not the TPW rate $\dot{\Theta}$)
in terms of $\theta$, $\alpha$, and $\beta$:
\begin{equation}
\dot{\theta} = 
-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\alpha \sin\beta.
\label{eq:theta_competition}
\end{equation}
Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it arbitrarily via
relative motion between the geographic frame and the principal axes of the $\mathbf{E}$-frame.
Comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
then the rate of TPW $\vert \dot{\Theta} \vert$ is identical to $\vert \dot{\theta} \vert$.
Conversely, when the rate of TPW is very slow (i.e. $\vert \dot{\Theta} \vert \sim 0$), 
then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$ 
and $\theta$ is completely determined by $\mitbf{\Psi}$
(where we have neglected the orientation factor $\sin \alpha \sin \beta$ for the purposes of scaling).
In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 
at least occasionally, be larger than its decay.

>>>>

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/reference_frames.pdf}
\caption{Relevant vectors and angles for the TPW analysis. The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively. Since the choice of geographic axes is arbitrary, we assume that at this instant the $\mathbf{E}$-frame and the geographic frames are colocated (though at future times they will not be). The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $\mathbf{e}_3$-axis. For illustration, we assume the longitude $\psi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane. True polar wander in the geographic reference frame moves the rotation axis towards the $\mathbf{e}_3$-axis, shown by $\dot{\mitbf{\omega}}$. However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$-frame. This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$ in the geographic frame, which is defined by colatitude $\beta$ and longitude $\gamma$. The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$-frame. The motion of $\mitbf{\omega}$ in the $\mathbf{E}$-frame is shown by $\dot{\mitbf{\omega}}^\prime$. Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Sections~\ref{sec:reference_frames} and \ref{sec:theta}}
\label{fig:reference_frames}
\end{figure*}

@@

There are two processes at work which control the evolution of $\theta$, its growth through perturbations in the convective moment and 
its decay via relaxation of the pole towards the maximum moment of inertia.
We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
the rotation of the body-fixed frame relative to inertial space, as shown in Equation~\eqref{eq:euler}).
The $\mathbf{E}$-frame slowly rotates with respect to the geographic frame, described by the rotation vector $\mathbf{\Psi}$ 
(here $\mitbf{\Psi}$ defines the rotation of the $\mathbf{E}$-frame in the geographic frame, see Figure~\ref{fig:reference_frames}).
We can then write the rate of change of $\mitbf{\omega}$ in the $\mathbf{E}$-frame (denoted by primes)
\begin{equation}
\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
Equation~\eqref{eq:liouville_ode} then becomes
\begin{equation}
 \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\label{eq:liouville_ode_new_frame}
\end{equation}
The advantage of expressing TPW in the $\mathbf{E}$-frame is that the entire equation
may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that $\phi=0$
(for nonzero $\phi$ the equations become more complex, but the basic physics is unchanged).
The direction of $\Psi$ is given by colatitude $\beta$ and longitude $\gamma$, as shown in Figure~\ref{fig:reference_frames}.

We may solve Equation~\eqref{eq:liouville_ode_new_frame} for $\dot{\theta}$ 
(i.e., the colatitudinal velocity in the $\mathbf{E}$ frame, not the TPW rate $\dot{\Theta}$)
in terms of $\theta$, $\beta$, and $\gamma$:
\begin{equation}
\dot{\theta} = 
-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\beta \sin\gamma.
\label{eq:theta_competition}
\end{equation}
Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it arbitrarily via
relative motion between the geographic frame and the principal axes of the $\mathbf{E}$-frame.
Comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
then the rate of TPW $\vert \dot{\Theta} \vert$ is identical to $\vert \dot{\theta} \vert$.
Conversely, when the rate of TPW is very slow (i.e. $\vert \dot{\Theta} \vert \sim 0$), 
then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$ 
and $\theta$ is completely determined by $\mitbf{\Psi}$
(where we have neglected the orientation factor $\sin \beta \sin \gamma$ for the purposes of scaling).
In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 
at least occasionally, be larger than its decay.
